Semantic authoring in a flexiformal context --- Bulk annotation of rigorous documents

### 2025-05-09 doc/cicm25-paper/paper.tex
paragraph on scaling snify

\section{Tool Support for Semantic Authoring in \sTeX/\ALeA}\label{sec:tools}
Semantic authoring can ``inherit'' tool support from both informal and formal authoring.
In the case of \sTeX, we can re-use spell checkers for traditional informal authoring.
There also is prototypical support for authoring \ALeA content in Microsoft Word
with a custom plugin (see \cite{KohKoh:woide24}),
which allows authors to use Word's spell checker and grammar checker.

@@

\section{The \snify System}\label{sec:snify}
The \snify\footnote{
From ``\texttt{\textbackslash sn}-ify''; ``\texttt{\textbackslash sn}'' is
a frequently used \sTeX short-hand macro for term annotations.}
utility \cite{stextools:git}
is a simple command-line tool that creates a catalog of
symbol-verbalization pairs, analyzes \sTeX source files, and then steps through all word
occurrences in the document that match a verbalization in the catalog.  Matching is based
on word stems using off-the-shelf stemmers, which allows us to e.g.\ match ``utilities''
with ``utility''.  For each matched word, the user is presented with an annotation choice
and interactions that allow to fine-tune the wordwise annotation
workflow.
The interface is inspired by traditional spell checkers like \lstinline|ispell|.
\Cref{fig:snify} shows a typical situation.

@@

As a companion to \snify,
we have \defianno, a prototype tool for annotating definienda in documents
that were originally authored without semantic annotations.
It can be configured to use different \LaTeX\ macros
to identify definienda (e.g.\ \lstinline|\emph| or \lstinline|\textbf|),
and then steps through them similarly to \snify.
For each potential definiendum, the user can then use fuzzy search to find
the corresponding symbol in the domain model.
This populates the catalog with verbalizations relevant to the document,
making subsequent \snify runs more effective,
especially for documents in languages that are not well covered by the
domain model (i.e.\ not English).

@@

\begin{wrapfigure}r{5.5cm}
  \begin{tikzpicture}[scale=.7]
    \tikzstyle{doc}=[draw,thick,align=center,color=black,
                  shape=document,minimum width=10mm,minimum height=15mm]
    \tikzstyle{database}=[cylinder,shape border rotate=90,aspect=0.25,draw, 
     cylinder uses custom fill,cylinder body fill=yellow!30,cylinder end fill=yellow!30]
     \tikzstyle{include}=[right hook-angle 45,thick]
     \tikzstyle{includeleft}=[left hook-angle 45,thick]
    \node[doc] (cc) at (0,0) {\shortstack{Content\\ Commons}};
    \node[doc] (pl) at (5,0) {\shortstack{Private/\\Local}};
    \node[database] (db) at (3,2.5) {\shortstack{\tiny Verbalization\\Cache}};
    \node (sn) at (0,4.5) {\texttt{snify}};
    \node (id) at (2.5,5) {IDE};
    \node (wo) at (5,4.5) {WOIDE};
    \draw[include]  (cc) -- (db);
    \draw[includeleft] (pl) -- (db);
    \draw[->] (db) -- (sn);
    \draw[->] (db) -- (id);
    \draw[->] (db) -- (wo);
    \draw[dashed,thick] (3,1.6)  -- (3,-1.4);
  \end{tikzpicture}
  \caption{Architecture}\label{fig:arch}\vspace*{-2em}
\end{wrapfigure}
One practical problem with \snify that remains is that the system only harvests
symbol/verbalization pairs from \sTeX files on the local file system. This means that, for
effective annotation, users (have to) download the whole content commons and keep it
updated. Unfortunately, the current MMT-based MathHub knowledge management
API\ednote{intro this above} does not supply verbalization information. The iMMT-based
API~\cite{iMMT:on} currently under development will, so \snify will be able to request the
pairs for the remaining (non-local) content commons.\ednote{
    This is already outdated.
    What is the current API status?
    How is it related to FLAMS?
}
If a pair is chosen in an annotation,
\snify can then automatically download the corresponding archive so that the annotated
file can be formatted with pdflatex or rusTeX locally.

@@

% LocalWords:  tbw Aarne Ranta homostem pre hmmm Vre ar al sn ify TODO et Hutter Plivelic
% LocalWords:  lang

>>>>

\section{Tool Support for Semantic Authoring in \sTeX/\ALeA}\label{sec:tools}

Semantic authoring can ``inherit'' tool support from both informal and formal authoring.
In the case of \sTeX, we can re-use spell checkers for traditional informal authoring.
There also is prototypical support for authoring \ALeA content in Microsoft Word with a
custom plugin (see \cite{KohKoh:woide24}), which allows authors to use Word's spell
checker and grammar checker.\ednote{MK: Mention CPoint and SMD as well? Probably yes.}

@@

\section{The \snify System}\label{sec:snify}

The \snify\footnote{ From ``\texttt{\textbackslash sn}-ify''; ``\texttt{\textbackslash
    sn}'' is a frequently used \sTeX short-hand macro for term annotations.}  utility
\cite{stextools:git} is a simple command-line tool that creates a catalog of
symbol-verbalization pairs, analyzes \sTeX source files, and then steps through all word
occurrences in the document that match a verbalization in the catalog.  Matching is based
on word stems using off-the-shelf stemmers, which allows us to e.g.\ match ``utilities''
with ``utility''.  For each matched word, the user is presented with an annotation choice
and interactions that allow to fine-tune the wordwise annotation workflow.  The interface
is inspired by traditional spell checkers like \lstinline|ispell|.  \Cref{fig:snify} shows
a typical situation.

@@

\paragraph{Definition Handling as a Pre-process for Multilingual Documents}
As a companion to \snify, we have \defianno, a prototype tool for annotating definienda in
documents that were originally authored without semantic annotations.  It can be
configured to use different \LaTeX\ macros to identify definienda (e.g.\ \lstinline|\emph|
or \lstinline|\textbf|), and then steps through them similarly to \snify.  For each
potential definiendum, the user can then use fuzzy search to find the corresponding symbol
in the domain model.  This populates the catalog with verbalizations relevant to the
document, making subsequent \snify runs more effective, especially for documents in
languages that are not well covered by the domain model (i.e.\ not English).

\begin{newpart}{MK: rewritten as existing but experimental. This needs to be placed
    somewhere else.}
  \paragraph{Scaling \snify to the Flexiformal Content Commons }
One practical problem with \snify system as presented so far is that it only harvests
symbol/verbalization pairs from \sTeX files on the local file system. This means that, for
effective annotation, users (have to) download the whole content commons and keep it
updated -- an assumption/requirement that becomes increasingly impractical as the \sTeX
ecosystem grows.

The general situation is that on the one hand there is a constantly growing, shared
\textbf{content commons} of public math archives -- currently 200 with about 14.000 \sTeX
files -- that are regularly converted to FTML (\underline{F}lexiformal \underline{T}ext
\underline{M}arkup \underline{L}anguage) and served on \url{https://MathHub.info}. On the
other hand authors are working on
\begin{compactenum}[\em i\rm)]
\item a set of local working copies of the archives -- with the intention of publishing
  them on MathHub.info and their dependencies (they need to be local, since pdflatex
  cannot deal with remote files) and
\item possibly a set of private archive (think papers under development or exam
  problems/solutions) 
\end{compactenum}
This situation is familiar from software development and also from e.g. the \LaTeX
ecosystem. Semantic authoring -- and thus the \snify utility -- inherits the requirements
from both.

\begin{wrapfigure}r{5.5cm}\vspace*{-2em}
  \begin{tikzpicture}[scale=.7]
    \tikzstyle{doc}=[draw,thick,align=center,color=black,
                  shape=document,minimum width=10mm,minimum height=15mm]
    \tikzstyle{database}=[cylinder,shape border rotate=90,aspect=0.25,draw, 
     cylinder uses custom fill,cylinder body fill=yellow!30,cylinder end fill=yellow!30]
     \tikzstyle{include}=[right hook-angle 45,thick]
     \tikzstyle{includeleft}=[left hook-angle 45,thick]
    \node[doc] (cc) at (0,-1) {\shortstack{Content\\ Commons}};
    \node[doc] (pl) at (5,-1) {\shortstack{Private/\\Local}};
    \node[database] (db) at (3,2.5) {\shortstack{\tiny Verbalization\\Cache}};
    \node[draw] (sn) at (0,4.5) {\snify};
    \node[draw] (id) at (2.5,5) {IDE};
    \node[draw] (wo) at (5,4.5) {WOIDE};
    \draw[draw,fill=black!10] (-1.3,.5) rectangle ++ (6.4,1); 
    \node (f) at (-.5,.9) {\flams};
    \draw[include]  (cc) -- node[left] {\tiny F2RDF} (db);
    \draw[includeleft] (pl) -- node[right] {\tiny linter} (db);
    \draw[->] (db) -- (sn);
    \draw[->] (db) -- (id);
    \draw[->] (db) -- (wo);
    \draw[dashed,thick] (3,1.6)  -- (3,-2);
    \draw[dashed,thick] (3,1.6)  -- (3,-2);
    \draw[includeleft] (pl) to[bend right=10] (wo);
  \end{tikzpicture}
  \caption{Architecture}\label{fig:arch}\vspace*{-1em}
\end{wrapfigure}
To cope with this situation, We have extended the brand-new \flams system
\cite{flams:on}\footnote{A paper on \flams is submitted to CICM 2025 and will be cited
  here if accepted.} with functionality for handling definienda in the FTML2RDF harvester,
so that all symbol/verbalizations are available in the \flams triple-store. Thus a harvest
from the flexiformal content commons is just a simple SPARQL query and some postprocessing
of the result (de-duplication, stemming, and pairing) away. Note that \flams works
directly on FTML, which is generated from \sTeX input by \textsf{rus\TeX} and exported
from annoated MS Word files via WOIDE.

In local/private archives, the \snify harvester needs to work directly on the \sTeX
sources. Here we have extended the \flams \textbf{linter} -- a simple \sTeX parser that
drives the LSP functionality the VSCode and \texttt{emacs} \sTeX IDE plugins use to deal
with definienda. This is much more efficient (and robust) than the initial regex-based
python parser). The analogon for this on the WOIDE side is still under development it
could be based on FTML generated by WOIDE, or WOIDE could export symbol/verbalization
pairs into the \snify verbalization cache directly.

In this situation, the main functionality of \snify is to maintain the verbalization cache
and answer front-end queries for verbalizations efficiently so that we can power various
front-ends from that. These include
\begin{compactenum}[\em i\rm)]
\item The traditional text/terminal-based \snify UI for bulk annotation,
\item the \sTeX IDEs in VSCode and \texttt{emacs} that can supply while-editing annotation
  support, and 
\item WOIDE, which can be seen as a semantic authoring IDE based on MS Word.\ednote{MK:
    CPoint and semantic wikis.}
\end{compactenum}

Note that implemented this broadly \snify can provide additionall logistic services: If a
pair is chosen in an annotation, \snify can then automatically download the corresponding
archive so that the annotated file can be formatted with \texttt{pdflatex} or
\textsf{rus\TeX} locally.
\end{newpart}

@@

@@

% LocalWords:  tbw Aarne Ranta homostem pre hmmm Vre ar al sn ify TODO et Hutter Plivelic
% LocalWords:  lang DHBKI FS Todo FTML lexiformal arkup anguage CICM RDF postprocessing
% LocalWords:  de
